The Natural History of Child Signals of Need in Utila, Honduras: An Exploratory Study

### 2024-11-06 Paper.qmd
Commiting final changes before circulating to co-authors.

---
title: "The Natural History of Child Signals of Need in Utila, Honduras: An Exploratory Study"
author:
  - name: Michael R. Gaffney
    orcid: 0000-0002-7552-6942
    corresponding: false
    affiliation:
      - name: Washington State University
        department: Anthropology
  - name: Jessica K. Hlay
    orcid: 0000-0002-6139-5144
    corresponding: false
    affiliation:
      - name: Boston University
        department: Anthropology
  - name: Izabel Rodriguez
  - name: Kristen L. Syme
    orcid: 0000-0002-5601-8897
    corresponding: false
    affiliation:
      - name: Leiden University
        department: Institute of Security and Global Affairs
  - name: Aaron D. Blackwell
    orcid: 000-0002-5871-9865
    corresponding: false
    affiliation:
      - name: Washington State University
        department: Anthropology
  - name: Edward H. Hagen
    orcid: 0000-0002-5758-9093
    email: edhagen@wsu.edu
    corresponding: true
    affiliation:
      - name: Washington State University
        department: Anthropology
  - name: Carolyn R. Hodges-Simeon
    orcid: 0000-0002-0691-9080
    corresponding: false
    affiliation:
      - name: Boston University
        department: Anthropology
execute:
  echo: false
  warning: false
format:
  html:
    toc: true
    toc-expand: 2
    number-sections: true
lightbox: true
editor: visual
tbl-cap-location: bottom
bibliography: signaling.bib
crossref:
  custom:
    - kind: float
      reference-prefix: Figure S
      key: sfig
      space-before-numbering: false
    - kind: float
      reference-prefix: Table S
      key: stbl
      space-before-numbering: false
---

@@

To gain support, children use signals to communicate their desires to parents. Infant signals of need, particularly infant cries, have been extensively studied in a range of populations. However, the full range of potential child signals of need, which extend beyond cries, has never been investigated in children of all ages. To address this, we used exploratory analyses on caregiver reports of child signaling data from xxx families (xx children) on Utila, Honduras. We found that child signaling decreased with age and neighborhood quality and was highly correlated with the frequency of conflict between children and caretakers. Consistent with signaling theory, children who were perceived as needier within the household were more likely to receive more investment. They were also more likely to receive a pro-child response when the cause of their last instance of signaling involved discomfort, pain, injury, or, illness, but less likely to receive a beneficial response when it involved prior punishment. Our results show there is much potential to the further study of sadness, crying, and temper tantrums among older children using signaling theory.

@@

Compared to chimpanzee mothers, mothers in hunter-gatherer societies have shorter interbirth intervals and produce offspring that require many more years of parental care [@davison2021; @davison2022]. Unlike chimpanzee offspring, human offspring in ancestral populations therefore often grew up with multiple older and/or younger siblings who were potential competitors for parental investment and yet who could also provide (or require) allocare [@helfrecht2016; @lawson2016; @hagen2009].

@@

For the same reason, siblings also often find themselves in conflicts of interest with one another, with each being expected to prefer more resources for themself than others. Sibling competition is evident across taxa and can range from relatively benign forms like strong begging competition in chicks [@caro2016] to siblicide in birds, mammals, and sharks [@salmon2014]. In humans, sibling competition is largely estimated by counts of siblings, and large numbers of siblings have been associated with decreased growth [@hadley2011; @hagen2006; @salmon2014], survival [@lawson2012; @meij2009; @strassmann2011], and reproductive success [@gibson2011; @mace1996] in multiple populations, highlighting the likely significance of sibling competition for resources, even if this does not always manifest with behavioral conflict [@bateson1994].

@@

In order to elicit more support, children often use signals of needs and wants. Across the tree of life, signals evolve to reveal agents’ private information to other agents when doing so benefits both the signaler and the receiver [@caro2016; @maynard-smith2004; @preuschoft2000; @setchell2005; @zahavi1975]. Children often have needs and wants, where needs involve avoiding potential fitness costs, such as poor nutrition, injuries, or threats to status, and wants involve acquiring potential fitness benefits, such as highly nutritious foods and valuable artifacts. Needs and wants are often not apparent to caregivers, however, i.e., they are private information, and hence children must signal them to receive help. Signals differ from cues in that signals were favored by natural selection due to their ability to alter the behavior or perceptions of others in ways that benefit the signaler, on average, whereas cues are information organisms use to influence their behavior or physiological state that did not evolve for communicative effect [@maynardsmith1995]. Crying when injured is likely a signal of need[\[1\]](#_ftn1), for instance, whereas bleeding would be a cue of need.

@@

Although widely accepted as a signal, there is no consensus on why emotional crying might elicit positive responses. As the costs and benefits of signaling and gaining investment change with age, it is likely the strategic logic of crying changes too. It could be vital for newborns to signal health and vigor to reduce the risk of neglect or infanticide experienced by children with poor survival prospects [@daly1984; @soltis2004]. As crying comes with physiological costs [@rao1997; @thureen1998], it could serve as a credible signal of vigor if these costs are too high for unhealthy infants to bear [@furlow1997; @lummaa1998; @soltis2004; @wells2003]. Alternatively, crying might impose costs on caregivers by, e.g., attracting predators or interfering with important activities, motivating caregivers to invest more to stop the crying [so-called blackmail or bargaining models; @xx, @yy]. More generally, contact calls between parents and offspring have evolved in numerous species, so cries might simply be a means for infants to elicit attention from caregivers [@mehr2021]. For older children, the relatively insignificant energetic costs of crying might not be enough to ensure it is a credible signal when there are conflicts of interest. For this reason, crying among older children might be more similar to adult crying, which can increase support and empathy but sometimes also reduces reputation and increases stigmatization [@cornelius2001; @cretser1982; @hendriks2006; @jesser1989; @vingerhoets2016]. Such costs could be a means by which crying can still be a credible signal of need if they dissuade potential fakers from making the signal [@sznycer2023] when there are moderate conflicts of interest.

@@

In all forms of recruitment, adult participants were informed of the purpose of the study, what they would be asked to do, and the inclusion criteria, such as being 18 or older, having permanent residence on Utila, and having a child between 6 and 18 years old.

\[no longer needed\] XXXXXXXXCHIld RecruitmentXXXXXXXXXXX

@@

All participants for the survey portion of the study met the inclusion criteria. Most ~~adult~~ participants were the biological mothers of one or more children in the study. ~~All children in the study, whose ages ranged from 5-20, were cared for by one of the adult participants.~~

@@

All ~~adult~~ participants provided informed consent. The surveys, which included many topics outside of child signaling, were designed to last about an hour, plus 20 minutes for each additional child. ~~Adults~~ Participants were paid \$XXXXX for the survey portion of the study.

~~xxxxxxxxxxxxxChild consent infoxxxxxxxxxxxxxx.~~

This study was approved by the xxxxxxxIRBINFOxxxxxxxxxxxxx.

@@

|  |  |
|-------------------------|-----------------------------------------------|
| Variable | Survey question wording |
| Crying Frequency | How often does your child cry? |
| Sad Frequency | How often is your child sad? |
| Tantrum Frequency | How often does your child throw temper tantrums? (e.g., scream, throws things, or throws themselves on the floor) |
| Running Away Frequency | How often does your child leave the house without telling you when they get mad? |
| Parent/Child Conflict Frequency | How often do you have conflicts with your child? |
| Child Alloparenting Frequency | How often does your child take care of their other siblings? |

@@

Frequencies of caregiver response to the last instance of child signaling by the cause of the signaling. Categories of causes were based on cluster analysis of more fine-grained cause variables. See @sfig-cause-clustering for cluster dendrogram and @sfig-sumtab-causes for the frequencies of causes and caregiver responses.
:::

@@

Caregivers reported a wide range of child signaling frequencies. About half of children were reported to cry or throw tantrums more than once a month and almost 40% were reported to do so more than once a week [@fig-barplot_SignalFreq]. Frequencies of sadness, crying, and tantrums were positively correlated with each other and with ~~family conflict~~ the frequency of conflict with the primary caretaker [@fig-signal_corrplot]. They also loaded on PC1 in the same direction as indices of adversity, such as food insecurity, and in the opposite direction of indices of wealth and environmental quality, such as possessions, neighborhood quality, and income [@fig-PCA]. Signaling decreased with child age, and increased with ~~family~~ conflict with their primary caretaker, which were the two variables most closely linked with signaling in the MST [@fig-mst]. In our lasso regression models (which excluded family conflict due to the likely bidirectional relationship between signaling and perceived conflict), child age and high neighborhood quality were consistently negatively associated with signaling frequency, as were household income, caregiver education, and number of adults helping with childcare. Medical problems were a consistent, albeit small positive predictor [@fig-lasso; @fig-signal-effects].

These results are consistent with the hypothesis that sadness, crying, and tantrums are signals of need, and that children's need is higher in adverse environments and at younger ages. One possible reason for the negative association of signaling with age, which is also seen in more industrialized settings [@zeifman2013; @potegal2003; @koch2003; @macfarlane1962], is that as children mature, they are more capable of fulfilling their own needs, more able to ask others for what they need [@osterman2010], and less vulnerable to threats. The effects of age on signaling are also likely due to how age influences the costliness of signaling need. For example, while the energetic cost of crying might be costly for the youngest children, it is likely insignificant for older children. Older children might instead need to weigh the reputational harm from crying against the potential increase in social support.

Consistent with signaling models that emphasize the role of conflicts of interest in favoring the use of costlier signals, caregiver perceptions of conflict and child signaling frequency were correlated [@fig-signal_corrplot], and family conflict was the most common cause of a child's recent signaling episode [@fig-effect_last_signal_all_responses]. Although conflicts of interest do not always lead to overt behavioral conflict [@altmann1980; @bateson1994], it is likely many, if not most, conflicts are ultimately caused by conflicts of interest. Conflicts, in turn, were negatively associated with child age and higher neigborhood quality, and positively associated with medical problems and (oddly) caregiver education [@fig-conflict-effects].

Caregiver responses to signals were generally positive or neutral, more so for situations involving injury and illness, and less so for situations involving family conflict, loss of privileges, and transgressions [@fig-effect_last_signal_all_responses; @fig-effect_last_signal_response]. These patterns probably reflect a combination of confluences of interest vs. conflicts of interest between child and caretaker, and less caretaker uncertainty about need in cases of injury and illness. That said, our model predicts positive responses in `r #signif(updated_out$estimate[3] * 100, 2)`% of cases involving trangressions (when all other predictors are held at their mean values), suggesting that children are successful in negotiating better outcomes at least some of the time.

@@

Interestingly, older children were perceived to be needier than younger ones [@@fig-effect_relativeneed]. One likely reason for this might be that caretakers perceive the types of investment older children seek as more costly. In line with this reasoning, multiple caretakers mentioned that older kids need more money and advice and get into trouble more often. It might also be that caretakers see investing in older children as relatively costly as they seek to withdraw investment in order to invest in younger children.

<!- If this is the case, the positive correlation between signaling effort and conflict frequency may represent children's attempts at credibly signaling their value of additional investment or the costliness of punishment [@syme2019] to parents who would have good reason to doubt the trustworthiness of simple verbal requests (or the trustworthiness of facial sadness and crying for those resorting to temper tantrums). It is also likely that parents find some forms of child signaling aversive, and therefore see it as conflict in and of itself. ->

\<!- In times when an older child's valuation of potential increased support is small, such costs may be prohibitive. However, in times when a child places more value on potential support, those same costs could allow for a child to credibly signal need if they are too high to be worth it for deceptive signaling but still made up for by the benefits for credible signalers. -\>

Finally, based on previous research showing that family demands, such as caring for younger siblings, was an important source of parent-offspring conflict [@syme2019], we investigated the relationship between the frequency of alloparenting by children and their frequency of signaling. We found that higher frequencies of alloparenting by girls was associated with a higher frequency of tantrums and overall signaling costs, but this relationship was slightly negative for boys [@fig-alloparentingXsex]. In many societies, girls do more alloparenting than boys [@helfrecht2016; @weisner1977; @barry1971; @turke1988], which matches reports from our focus groups of Utilian mothers. It is plausible that older girls are pressured to do more care of their younger siblings, which in other studies was found to spark parent-offspring conflict [@lowe2003; syme2023a]. If so, girls with excessive alloparenting responsibilities might be signaling their need for more time to invest in peer relationships and status seeking behaviors outside the household. Opposed to this interpretation, our study found a small negative assocation between girls' alloparenting and caregiver reports of conflict [@fig-lasso], although perhaps when girls do what caregivers want, caregivers report less conflict.

--- Hagen stopped here ---

\<!- We found evidence of an interaction between how often a child cared for their siblings and the sex of the child in our lasso models examining signal frequencies and cost, whereby there was a positive relationship between the frequency and summed costliness of child signals of need and alloparenting effort for girls but no relationship or a slightly negative one for boys. -\>

<!--Evidence for conflicts similar to the one hypothesized here has been reported by multiple researchers working with the Chuukese. Lowe found that both male and female children reported conflict with their families over their desires to invest financial resources in their own status enhancing opportunities rather than in the household [@lowe2003]. In an analysis of conflict among Chuukese-Americans, Syme and Hagen found that severe conflicts were more likely with greater numbers of siblings and when conflicts involved a child spending time away from home or evading family responsibilities [@syme2023a], findings which would be consistent with parents pressuring older children into investing more in the household among other explanations. In response to such conflicts, participants reported using a wide range strategies as children including, deception, withholding cooperation, running away, and seeking the assistance of others. Through these means, participants were often able to resolve conflicts in their favor despite their lower status in the household and community.-->

<!--However, child alloparenting effort was negatively associated with caregiver reports of child conflict. One potential reason for this is since parents presumably value the child alloparenting within the family, they see their relationships with children who invest more in the household as less conflictual. In line with this reasoning, we found evidence of a child alloparenting effort by child age interaction, with our model predicting increased levels of conflict for older children with greater alloparenting \[if we say this we should report the results of that lasso too\]. Ideally, we could control for signaling behavior to test this. However, the bidirectional relationship between child signaling and conflict with caretakers prevents this. It is also possible that our hypothesis about conflict over investment in the household is incorrect.-->

<!--There was a positive association between the frequency of child sadness and caretaker perceptions of child neediness within the family. However, the other signaling measures examined had coefficients set to 0 in the same lasso model despite the expectation that their greater costliness would make them more credible sources of information to parents. One reason for this might be that costs of sadness, while minor short-term, can be significant long-term. These costs are clearly seen in depression, which often involves an extended period of a lack of interest in one's normal activities and decreased investment in oneself and one's cooperative relationships and has been suggested to capable of credibly signal need in times of substantial conflicts of interest [@watson2002; @hagen2003].-->

<!--Another possibility is that parents are using cues from sadness (i.e., withdrawal and rumination) to evaluate the relative neediness of children. As cues are not created for the purpose of communication, they are completely reliable if perceived and interpreted correctly. Therefore, if some cues are hard to or are impossible to fake, they may be more reliable than most signals. That said, it is likely that some symptoms of sadness are simultaneously the result of both signals and cues, something which may have been favored by natural selection if the threat of deception meant cue-level symptoms were insufficient in eliciting support without upregulation, something which has been suggested to be the case with various illness symptoms [@tiokhin2016; @steinkopf2015].-->

<!--Caretakers perceived older children as needier, despite having more conflicts with younger children. One likely reason for this might be that caretakers perceive the types of investment older children seek as more costly. In line with this reasoning, multiple caretakers mentioned that older kids need more money and advice and get into trouble more often during surveys. It may also be that caretakers see investing in older children as relatively costly as they seek to withdraw investment in order to invest in younger children.-->

<!--Signaling behavior did not predict differences in relative caregiver investment between children within households, beyond its effect on perceived need [@sfig-plot-invest-coefs]. However, greater perceptions of need among caregivers were associated with increased likelihood of a child being reported to receive more investment in terms of time, effort, and resources relative to other children within the household. As previously mentioned, credible signals are capable of leading to pro-signaler responses due to their ability to reveal private information to others who would be motivated to increase their investment in another should they have that information. In other words, much of the effect of signaling on levels of support is expected to be mediated by the change in the beliefs of others that one could benefit from increased support that it causes (e.g., that a child is truly in need due to adversity) rather than direct effects. In line with this reasoning, an experimental vignette study which found that depression and suicidality increased the likelihood of costly support from participants in times of uncertainty and substantial conflicts of interest also found that between 67% and 78% of the effect of signal type on the likelihood of costly support was mediated by the effect of signaling on belief in need [@gaffney2022].-->

<!--Parents were more likely to respond positively and less likely to respond negatively to the last time a child signaled when the cause involved discomfort, pain, injury, or illness than when it did not. One potential causal explanation for this is that parents may often have information for sources outside of child signaling to corroborate a child's need (e.g., a fever, a diagnosis, or witness a child hurt themselves) so that there is less private information about a child's current state. Simultaneously, conflicts of interest may be less significant in these times due to either 1) both sides benefiting similarly from their child returning to normal (e.g., with an illness) or 2) the cause involving something which is minor and would quickly resolve itself (e.g. a skinned knee). It may also be that many of the typical positive responses to pain and illness are not very costly to caretakers (e.g., providing comfort), thereby increasing their frequency.-->

<--When the cause involved the child having made a transgression, punishment, or the loss of privleges or an item, however, parents were more likely to respond negatively and less likely to respond positively. In these circumstances, children may be signaling about the severity of a punishment [@syme2019]. However, parents may be reluctant to reduce punishments if the behavior that warranted the punishment was repeated or severe. They may also react with scolding or further punishment. That said, our model predicts positive responses in `r #signif(updated_out$estimate[3] * 100, 2)`% of cases (when all other predictors are held at their mean values), suggesting that children are successful in negotiating better outcomes at least some of the time.-->

Parents were also more likely to respond negatively and less likely to respond positively to children when the cause involved family conflict. This might reflect tradeoffs within the family, whereby investment in one child cannot go to another. Caretakers may have also responded more negatively when the cause involved family conflict, if it involved a child imposing costs on other family members.

\[delete this paragraph?\] We found a negative association between caretakers’ education level (years in school) and many of our child signaling measures despite our models controlling for household income and partner status. One possible reason for this may be that more educated caretakers see their children less during the day due to greater ability to afford childcare or due to them having jobs that are less compatible with child care (test model of social support predicted by education). Unfortunately, we do not have data on how much time caretakers spent with children to test this, and it seems equally likely that less educated parents may see their kids less often if they need to work more hours to support their families. Another possibility is that lower levels of education are associated with greater levels of adversity for both parents and children, something which appears to be the case for other measures of SES [@pepper2017] and would increase the benefits of signaling need for children and the difficulty of parents to respond. It is also possible that education influences parenting norms in ways that cause differences in child signaling behavior.

Outside of the sex by alloparenting effort interaction, we found little evidence of sex differences in child signaling behavior. As adults, females cry more than males, a finding that has been reported across cultures [@vanhemert2011; @becht2001]. One reason why our study may have failed to find significant effects of sex is that sex differences in the causes and reputational effects of signaling need are small or nonexistent until adolescence and young adulthood, with our small sample size for this age range preventing us from detecting a real effect should it exist. It is also possible that the sex difference seen in rates of crying in adults may reflect the lesser ability of females to bargain for better treatment through aggression or the threat of it when faced with adversity [@hagen2016]. Such a sex difference may not exist among preadolescencents due to a lack of difference in strength or athletic ability before puberty [@handelsman2018] or due to their relative lack of formidability compared to adults.

Despite the differences in wealth, job certainty, discrimination, and living conditions [@davis2014; @garcia2017; @garcia2018; @stonich2000] between recent immigrants from the mainland and native Utilians, primary language, immigration status, and residence in Camponado did not predict child signaling in our lasso regressions. However, neighborhood quality, which was strongly negatively correlated with residence in Camponado, (check others) was preserved in the lasso regressions predicting every signaling measure but sadness and was among the variables with the largest negative coefficients. Monthly income was also negatively associated with crying frequency,\~\~ tantrum frequency (too small of a coefficient to mention here?),\~\~ summed signal frequency, and signal cost albeit with smaller coefficients.

Many models designed to explain signals of need (e.g., the begging of young birds) predict that offspring who are worse off or are receiving less investment have more to gain from signaling than those with more [@mock2011; @godfray1991; @caro2016a; @godfray2000] and multiple game theoretical models examining parent-offspring conflict have found that parental resources have a modest impact on the evolutionary stable value for offspring demand for resources [@bossan2013; @mock1997 [confirm]]. In line with this reasoning, children in lower income households and riskier environments might benefit more from signaling effort if they experience greater levels of adversity or perceive themselves as disadvantaged compared to other children they may be competing with for status and desirable cooperative relationships. ~~In addition to overall effort in eliciting need, resource availability may have also affect how children signal for more investment based on the finding that birds in stable and unusually favorable environments are more likely to compete with siblings through signals thought to reveal information about need, while those in relatively unpredictable and unusually poor environments have been found to compete more through signals thought to reveal information about quality [@caro2016a].~~

(temporarilly moved here) Tantrums were present among older children and might be an under-appreciated way by which children signal for increased support. Most studies of tantrums involve children under 4 years old or older children with disruptive mood dysregulation disorder, likely due to their greater prevalence in these populations (cite). However, multiple studies which included older children from non-clinical populations have found similar evidence of tantrums [@macfarlane1962; @barbarin1999; @ward2013; @osterman2010; @goodenough1931; @bhatia1990], as have ethnographic studies [@ward2013; @lessa1950; @dennis1940].

@@

<!--Due to the fact the data used here were a part of survey which covered many topics and often multiple children, we also faced a tradeoff between survey time and question depth. Asking adult participants about their relative investment in children in their household was likely not as effective as more in depth, questions focused on specific questions. Framing the question this way may have also increased the likelihood of social desirability bias [@grimm2010], which could lead adult participants to be be more likely to report equal investment across children. Our questions about the last time the child signaled may have also been more effective if they focused on specific signals due to our methods combining crying, which is unambiguous as a signal, with sadness, which has both signaling and non-signaling functions. -->

@@

**Footnotes**

[\[1\]](#_ftnref1) We expect “signals of need” to actually be signals of resource valuation. However, we use “signal of need” here to distinguish signals that involve attempts at eliciting support through revealing private information about one’s vulnerabilities from the more commonly studied signals which reveal private information about one’s positive qualities (i.e., those related to partner choice and intrasexual competition).

[\[2\]](#_ftnref2) Our use of “credible” signaling is largely synonymous with the more common term, honest signaling. However, we prefer the term credible signals since signals do not need to be completely honest for natural selection to have favored individuals who paid attention to them.

[\[3\]](#_ftnref3) Researchers often distinguish between sadness and low mood based on the length of negative affect. We remain agnostic on how different sadness and low mood are from one another but use the terms researchers used when describing their research.

[\[4\]](#_ftnref4) All else is often not equal when it comes to the costliness of investment as older children may often cost more (e.g., the costs education and food are often higher among older children). However, parents would still have to pay similar costs to younger children as they grew, something which natural selection has likely acted on when it comes to our adaptations involved with differential child investment.

[\[5\]](#_ftnref5) Even in instances where conflicts begin due to mistakes, the initial conflict can lead to increased conflicts of interest. This may be seen with findings that individuals who accidentally harm another often grow to dislike them (cite), something which has been suggested to be adaptive to the accidental perpetrator as the unintended victim is likely to value than they would have before, potentially to the point of wishing them harm.

[\[6\]](#_ftnref6) Despite the risk, Shostak notes there were no known recent incidents of harm to women who ran away to spend the night in the Bush [@shostak1981].

>>>>

---
title: "The Natural History of Child Signals of Need in Utila, Honduras: An Exploratory Study"
author:
  - name: Michael R. Gaffney
    orcid: 0000-0002-7552-6942
    corresponding: false
    affiliation:
      - name: Washington State University
        department: Anthropology
  - name: Jessica K. Hlay
    orcid: 0000-0002-6139-5144
    corresponding: false
    affiliation:
      - name: Boston University
        department: Anthropology
  - name: Izabel Rodriguez
  - name: Kristen L. Syme
    orcid: 0000-0002-5601-8897
    corresponding: false
    affiliation:
      - name: Leiden University
        department: Institute of Security and Global Affairs
  - name: Aaron D. Blackwell
    orcid: 000-0002-5871-9865
    corresponding: false
    affiliation:
      - name: Washington State University
        department: Anthropology
  - name: Edward H. Hagen
    orcid: 0000-0002-5758-9093
    email: edhagen@wsu.edu
    corresponding: true
    affiliation:
      - name: Washington State University
        department: Anthropology
  - name: Carolyn R. Hodges-Simeon
    orcid: 0000-0002-0691-9080
    corresponding: false
    affiliation:
      - name: Boston University
        department: Anthropology
execute:
  echo: false
  warning: false
format:
  docx:
    number-sections: true
  html:
    toc: true
    toc-expand: 2
    number-sections: true
lightbox: true
editor: visual
tbl-cap-location: bottom
bibliography: signaling.bib
crossref:
  custom:
    - kind: float
      reference-prefix: Figure S
      key: sfig
      space-before-numbering: false
    - kind: float
      reference-prefix: Table S
      key: stbl
      space-before-numbering: false
---

@@

To gain support, children use signals to communicate their needs and wants to parents. Infant signals of need, particularly infant cries, have been extensively studied in a range of populations. However, the full range of potential child signals of need, which extend beyond cries, has rarely been investigated in a single study of children of all ages. To fill this gap, we collected mother and other primary caregiver reports of child signaling from `{r} nrow(caregivers)` families with `{r} nrow(modeldf)` children on Utila, a small island off the coast of Honduras. In exploratory analyses, we found that child signaling was common in both sexes and across all ages, although it decreased with age and neighborhood quality and increased with the frequency of conflict between children and caretakers. Consistent with signaling theory, children who were sad more frequently were perceived as needier within the household and were more likely to receive investment. Consistent with previous research, increased alloparenting effort was associated with more signaling by girls but less signaling by boys. Our results suggest that evolutionary theories of signaling can help explain patterns of child sadness, crying, and temper tantrums.

@@

Compared to chimpanzee mothers, mothers in hunter-gatherer societies have shorter interbirth intervals and produce offspring that require many more years of parental care [@davison2021; @davison2022]. Unlike chimpanzee offspring, human offspring in ancestral populations therefore often grew up with multiple older and/or younger siblings who were potential competitors for parental investment and yet who could also provide (or require) allocare [@helfrecht2016; @lawson2016; @hagen2009]. Juveniles in humans, unlike chimps, therefore evolved in families with multiple siblings simultaneously dependent on parents and others for essential care, likely selecting for credible signals of need.

@@

For the same reason, siblings also often find themselves in conflicts of interest with one another, with each expected to prefer more resources for himself or herself than others. Sibling competition is evident across taxa and can range from relatively benign forms like strong begging competition in chicks [@caro2016] to siblicide in birds, mammals, and sharks [@salmon2014]. In humans, sibling competition is largely estimated by counts of siblings, and large numbers of siblings have been associated with decreased growth [@hadley2011; @hagen2006; @salmon2014], survival [@lawson2012; @meij2009; @strassmann2011], and reproductive success [@gibson2011; @mace1996] in multiple populations, highlighting the likely significance of sibling competition for resources, even if this does not always manifest with behavioral conflict [@bateson1994].

@@

In order to elicit more support, children often use signals of needs and wants. Across the tree of life, signals evolve to reveal agents’ private information to other agents when doing so benefits both the signaler and the receiver [@caro2016; @maynard-smith2004; @preuschoft2000; @setchell2005; @zahavi1975]. Children often have needs and wants, where needs involve avoiding potential fitness costs, such as poor nutrition, injuries, or threats to status, and wants involve acquiring potential fitness benefits, such as highly nutritious foods and valuable artifacts. Needs and wants are often not apparent to caregivers, however, i.e., they are private information, and hence children must signal them to receive help. Signals differ from cues in that signals were favored by natural selection due to their ability to alter the behavior or perceptions of others in ways that benefit the signaler, on average, whereas cues are information organisms use to influence their behavior or physiological state that did not evolve for communicative effect [@maynardsmith1995]. Crying when injured is likely a signal of need, for instance, whereas bleeding would be a cue of need.

@@

Although widely accepted as a signal, there is no consensus on why emotional crying might elicit positive responses. As the costs and benefits of signaling and gaining investment change with age, it is likely the strategic logic of crying changes too. It could be vital for newborns to signal health and vigor to reduce the risk of neglect or infanticide experienced by children with poor survival prospects [@daly1984; @soltis2004]. As crying comes with physiological costs [@rao1997; @thureen1998], it could serve as a credible signal of vigor if these costs are too high for unhealthy infants to bear [@furlow1997; @lummaa1998; @soltis2004; @wells2003]. Alternatively, crying might impose costs on caregivers by, e.g., attracting predators or interfering with important activities, motivating caregivers to invest more to stop the crying [so-called blackmail or bargaining models; @zahavi1977a; @eshel1991; @bergstrom1999 @kennedy2021]. More generally, contact calls between parents and offspring have evolved in numerous species, so cries might simply be a means for infants to elicit attention from caregivers [@mehr2021]. For older children, the relatively insignificant energetic costs of crying might not be enough to ensure it is a credible signal when there are conflicts of interest. For this reason, crying among older children might be more similar to adult crying, which can increase support and empathy but sometimes also reduces reputation and increases stigmatization [@cornelius2001; @cretser1982; @hendriks2006; @jesser1989; @vingerhoets2016]. Such costs could be a means by which crying can still be a credible signal of need if they dissuade potential fakers from making the signal [@sznycer2023] when there are moderate conflicts of interest.

@@

In all forms of recruitment, participants were informed of the purpose of the study, what they would be asked to do, and the inclusion criteria, such as being 18 or older, having permanent residence on Utila, and having a child between 6 and 18 years old.

@@

All participants for the survey portion of the study met the inclusion criteria. Most participants were the biological mothers of one or more children in the study.

@@

All participants provided informed consent. The surveys, which included many topics outside of child signaling, were designed to last about an hour, plus 20 minutes for each additional child. Participants were paid \$XXXXX for the survey portion of the study.

This study was approved by the WSU xxxIRBINFOxxx and BU xxxIRBINFOxxx.

@@

|  |  |
|--------------------------|----------------------------------------------|
| Variable | Survey question wording |
| Crying Frequency | How often does your child cry? |
| Sad Frequency | How often is your child sad? |
| Tantrum Frequency | How often does your child throw temper tantrums? (e.g., scream, throws things, or throws themselves on the floor) |
| Running Away Frequency | How often does your child leave the house without telling you when they get mad? |
| Parent/Child Conflict Frequency | How often do you have conflicts with your child? |
| Child Alloparenting Frequency | How often does your child take care of their other siblings? |

@@

Frequencies of caregiver response to the last instance of child signaling by the cause of the signaling. Categories of causes were based on cluster analysis of more fine-grained cause variables. See @sfig-cause-clustering for cluster dendrogram and @stbl-sumtab-causes for the frequencies of causes and caregiver responses.
:::

@@

Caregivers reported a wide range of child signaling frequencies. About half of children were reported to cry or throw tantrums more than once a month, and almost 40% were reported to do so more than once a week [@fig-barplot_SignalFreq]. Frequencies of sadness, crying, and tantrums were positively correlated with each other and with the frequency of conflict with the primary caretaker [@fig-signal_corrplot]. They also loaded on PC1 in the same direction as indices of adversity, such as food insecurity, and in the opposite direction of indices of wealth and environmental quality, such as possessions, neighborhood quality, and income [@fig-PCA].

Signaling decreased with child age, and increased with conflict with their primary caretaker, which were the two variables most closely linked with signaling in the MST [@fig-mst]. In our lasso regression models (which excluded family conflict due to the likely bidirectional relationship between signaling and perceived conflict), child age and high neighborhood quality were consistently negatively associated with signaling frequency, as were household income, caregiver education, and number of adults helping with childcare. Medical problems were a consistent, albeit small positive predictor [@fig-lasso; @fig-signal-effects].

Despite the differences in wealth, job certainty, discrimination, and living conditions between recent immigrants from the mainland and native Utilians [@davis2014; @garcia2017; @garcia2018; @stonich2000], primary language, immigration status, and residence in Camponado did not predict child signaling in our lasso regressions. These variables correlated with neighborhood quality, however, which was retained in our models, and lasso regression tends to retain only one of a set of correlated predictors [@zou2005].

The foregoing results are consistent with the hypothesis that sadness, crying, and tantrums are signals of need, and that children's need is higher in adverse environments and at younger ages. One possible reason for the negative association of signaling with age, which is also seen in more industrialized settings [@zeifman2013; @potegal2003; @koch2003; @macfarlane1962], is that as children mature, they are more capable of fulfilling their own needs, more able to ask others for what they need [@osterman2010], and less vulnerable to threats. The effects of age on signaling are also likely due to how age influences the costliness of signaling need. For example, while the energetic cost of crying might be costly for the youngest children, it is likely insignificant for older children. Older children might instead need to weigh the reputational harm from crying against the potential increase in social support.

Consistent with signaling models that emphasize the role of conflicts of interest in favoring the use of costlier signals, caregiver perceptions of conflict and child signaling frequency were correlated [@fig-signal_corrplot], and family conflict was the most common cause of a child's recent signaling episode [@fig-effect_last_signal_all_responses]. Although conflicts of interest do not always lead to overt behavioral conflict [@altmann1980; @bateson1994], it is likely many, if not most, conflicts are ultimately caused by conflicts of interest. Conflicts, in turn, were negatively associated with child age and higher neigborhood quality, and positively associated with male sex, medical problems, and (oddly) caregiver education [@fig-conflict-effects].

Caregiver responses to signals were generally positive or neutral, more so for situations involving injury and illness, and less so for situations involving family conflict, loss of privileges, and transgressions [@fig-effect_last_signal_all_responses; @fig-effect_last_signal_response]. These patterns probably reflect a combination of confluences of interest vs. conflicts of interest between child and caretaker, some of which might represent caretaker investment in some children more than others, and less caretaker uncertainty about need in cases of injury and illness. That said, our model predicts positive responses in 35% of cases involving trangressions, suggesting that children are successful in negotiating better outcomes at least some of the time.

@@

Interestingly, older children were perceived to be needier than younger ones [\@ @fig-effect_relativeneed]. One likely reason for this might be that caretakers perceive the types of investment older children seek as more costly. In line with this reasoning, multiple caretakers mentioned that older kids need more money and advice and get into trouble more often. It might also be that caretakers see investing in older children as relatively costly as they seek to redirect investment toward younger, needier children.

Finally, based on previous research showing that family demands, such as caring for younger siblings, was an important source of parent-offspring conflict [@syme2019], we investigated the relationship between the frequency of alloparenting by children and their frequency of signaling. Although there was little evidence for sex differences in signaling overall (perhaps because sex differences typically emerge in adolescence, and our sample of adolescents was small), there was evidence that sex interacted with alloparenting effort. We found that higher frequencies of alloparenting by girls was associated with a higher frequency of tantrums and overall signaling costs, whereas for boys this relationship was slightly negative [@fig-alloparentingXsex]. In many societies, girls do more alloparenting than boys [@helfrecht2016; @weisner1977; @barry1971; @turke1988], which matches reports from our focus groups of Utilian mothers. It is plausible that older girls are pressured to do more care of their younger siblings, which in other studies was found to spark parent-offspring conflict [@lowe2003; syme2023a]. If so, girls with excessive alloparenting responsibilities might be signaling their need for more time to invest in peer relationships and status seeking behaviors outside the household. Opposed to this interpretation, our study found a small negative assocation between girls' alloparenting and caregiver reports of conflict [@fig-lasso], although perhaps when girls do what caregivers want, caregivers report less conflict.

Many models designed to explain signals of need (e.g., the begging of young birds) predict that offspring who are worse off or are receiving less investment have more to gain from signaling than those with more [@mock2011; @godfray1991; @caro2016a; @godfray2000] and multiple game theoretical models examining parent-offspring conflict have found that parental resources have a modest impact on the evolutionary stable value for offspring demand for resources [@bossan2013; @mock1997 [confirm]]. In line with this reasoning, children in lower income households and riskier environments might benefit more from signaling effort if they experience greater levels of adversity or perceive themselves as disadvantaged compared to other children they may be competing with for resources, status, and valuable social relationships.

@@

@@
